Choosing Color Palettes for Statistical Graphics

### 2006-10-20 trunk/vcd/paper/Color/TAS.Rnw
draft almost finished, preparing submission version for TAS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@595 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

source("functions.R")
load("Bundestagswahl2005.rda")

@@

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is the \pkg{ColorBrewer.org} tool \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color palettes for maps, providing sets of colors
for coding various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables.
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors (with a limited
number of colors in each set) but describe paths through a perceptually-based color space
so that the user can decide which path exactly should be taken and how many colors should
be selected. The color space employed by us is the Hue-Chroma-Luminance (HCL) color space
\citep[see][]{vcd:Ihaka:2003} and we derive explicit formulae describing the paths through
this space that correspond to different types of palettes. Given a mapping from HCL to RGB
coordinates, our formulas are extremely easy to implement and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

@@

\setkeys{Gin}{width=.6\textwidth}
\begin{figure}[p]
\begin{center}
<<seats, fig=TRUE, height=5, width=6>>=
par(mar = rep(0.8, 4))
pie(seats, clockwise = TRUE, col = hcl(c(240, 60, 300, 120, 0), 60, 75), radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<votes, fig=TRUE, height=9, width=11>>=
Votes <- as.table(as.matrix(votes[c(1, 3, 4, 5, 9, 11, 13, 14, 15, 16, 2, 6, 7, 8, 10, 12),c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]))
names(dimnames(Votes)) <- c("", "")
dimnames(Votes)[[2]][c(4, 5)] <- c("Gr", "Li")

mosaic(Votes, labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center", "center", "center", "right")),
  gp = gpar(fill = hcl(c(240, 60, 0, 120, 300), 60, 75)),
  spacing = spacing_highlighting, labeling = labeling_left,
  margins = unit(c(2.5, 1, 1, 12), "lines"), keep = FALSE)
@
\caption{\label{fig:votes} Votes in the German election 2005.}
\end{center}
\end{figure}

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU better in the South; furthermore,
Die Linke performed particularly well in the Eastern provinces and in Saarland.

>>>>

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien, Austria}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

source("functions.R")

@@

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color palettes for maps, providing sets of colors
for coding various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables.
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors (with a limited
number of colors in each set) but describe paths through a perceptually-based color space
so that the user can decide which path exactly should be taken and how many colors should
be selected. The color space employed by us is the Hue-Chroma-Luminance (HCL) color space
\citep[see][]{vcd:Ihaka:2003} and we derive explicit formulae describing the paths through
this space that correspond to different types of palettes. Given a mapping from HCL to RGB
coordinates, our formulas are extremely easy to implement and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

@@

\setkeys{Gin}{width=.6\textwidth}
\begin{figure}[p]
\begin{center}
<<seats, fig=TRUE, height=5, width=6>>=
par(mar = rep(0.8, 4))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
names(parties) <- names(seats)
pie(seats, clockwise = TRUE, col = parties, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<votes, fig=TRUE, height=9, width=11>>=
data("Bundestag2005")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties <- parties[colnames(votes)]
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
mosaic(votes, gp = gpar(fill = parties),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE)
@
\caption{\label{fig:votes} Votes in the German election 2005.}
\end{center}
\end{figure}

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU/CSU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU/CSU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU/CSU better in the South; furthermore,
Die Linke performed particularly well in the Eastern provinces and in Saarland.
